Mining HIV Trends in Social Media Data

### 2016-05-05 draft/draft.tex
update draft

Pre-Exposure Prophylaxis (PrEP) is a recently developed therapy for the prevention of Human Immunodeficiency Virus (HIV) transmission. The treatment is available as a pill named Truvada and is taken once a day by HIV negative individuals for as much as 99\% protection when exposed to HIV. While medically effective, PrEP suffers from high drug costs, HIV-related social stigma, and under-informed health providers and patients. Data mining of social media has proven effective in the monitoring of HIV in the United States, but no study has investigated PrEP using social media. This paper describes a data mining and machine learning strategy using natural language processing (NLP) that monitors Twitter social media data to identify PrEP discussion trends. Our results show that we can identify PrEP and HIV discussion dynamics over time, and PrEP-related tweets with positive and negative sentiment. These results can be used by public health professionals to monitor PrEP discussion trends, and identify strategies to improve HIV prevention.

@@

Tweets from twitter consist of short 140 character text messages that may also contain hashtag annotations. Twitter has been used as a source of data for large scale opinion mining in public health monitoring contexts to predict the spread of influenza~\cite{aramaki2011twitter}, predict postpartum depression~\cite{de2013predicting}, and examine tobacco use~\cite{myslin2013using}. Recently it has also been used for the study of HIV~\cite{young2014methods,young2013online}. These studies of HIV have focused on county-level HIV prevalence prediction, and general HIV discussion monitoring, but they have not focused on PrEP related discussion. Furthermore, existing HIV social media analyses have not taken full advantage of natural language processing (NLP) techniques to discover semantic information in unstructured text~\cite{young2015big}.

@@

Latent Dirichlet Allocation~\cite{blei2003latent} (LDA) is used to identify a small set of latent topics present in an unstructured corpus. LDA is a graphical model that generates documents from a set of latent topics. A topic is a probability distribution over words that captures a set of related words. LDA models are often inferred in practice using Bayesian inference either via collapsed Gibbs sampling or Variational Bayesian inference. Inspection of the resulting topics allows one to identify relevant terms and the context in which they occur in the corpus.

Dynamic Topic Modelling~\cite{blei2006dynamic} (DTM) is an extension of LDA that produces a seried of topic models over time. Briefly, the documents in the corpus are divided into several corpuses that are successive in time. LDA is performed on each corpus to extract a topic distribution. The posterior topic distribution from time $t_n$ is used as the prior for time $t_{n+1}$. This lets the DTM model determine a topic model for each time point that is dependent both on the set of tweets from that time point, and on the previous time point's topic distribution.

@@

% table 1 - w2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to word-vector PrEP}
\begin{tabular}{|l|c|} \hline
Related Word & Cosine Similarity to PrEP\\ \hline
truvada & 0.796666\\ \hline
DoingIt & 0.738141\\ \hline
WorldAIDSDay & 0.720910\\ \hline
NancyReagan & 0.717667\\ \hline
NBHAAD & 0.705061\\ \hline
ART & 0.704300\\ \hline
HLM2016AIDS & 0.698698\\ \hline
ICASA2015 & 0.693117\\ \hline
HIV & 0.692860\\ \hline
OneConversation & 0.688427\\ \hline
\hline\end{tabular}
\end{table}

@@

TODO: investigate those tweets, especially "Checkout why PrEP is hurting the cause"

User 711275699529764864 appears to be a twitter spam bot with no obvious connection to PrEP. Thus while we observe some noise, our results demonstrate a way to quickly identify and monitor the most viral tweets related to PrEP. As the discussion changes, public health professionals can use this information to quickly identify the most relevant viral sentiment in the online PrEP conversation.

% table 2 - d2v PrEP
\begin{table}
\centering
\caption{Cosine Similarity to doc-vector \#PrEP}
\begin{tabular}{|l|c|} \hline
Related hashtag/Tweet & Cosine Similarity to \#PrEP\\ \hline
\#lgbtmedia16 & 0.739128\\ \hline
\#hiv & 	0.727602 \\ \hline
\#whereisprep & 0.707165 \\ \hline
\#truvada & 0.696113 \\ \hline
\#hivprevention & 0.636068 \\ \hline
tweet-702179860983189504 & 0.630055\\ \hline
user-711275699529764864 & 0.629254\\ \hline
tweet-708519265540907010 & 0.628778 \\ \hline
tweet-712032637024653313 & 0.628646 \\ \hline
\#harrogatehour & 0.628547 \\ \hline
\hline\end{tabular}
\end{table}

@@

% do a top positive tweets table (also with tweet ID)
% table 3
\begin{table*}
\centering
\caption{Positive Sentiment Tweets.}
\begin{tabular}{|p{2.5cm}|p{12cm}|} \hline
Category & Text\\ \hline
General & "RT TOPublicHealth The Works provides testing for HIV anonymous \& rapid test available . Call 416-392-0520 for more info"\\ \hline
General & "RT FCAA ejaforg announced 5.4 million in grants to support orgs addressing \#HIV in new \& innovative ways!"\\ \hline
General & "RT HillaryClinton A note on the fight against HIV and AIDS and the people who really started the conversation."\\ \hline

@@

% do a top negative tweets table (also with tweet ID)
% table 4
\begin{table*}
\centering
\caption{Negative Sentiment Tweets.}
\begin{tabular}{|p{2.5cm}|p{12cm}|} \hline
Category & Text\\ \hline
General & "Also, how fucking vile of Hillary to say. Reagan did fucking NOTHING during the AIDS epidemic until it was too late. What a stupid old hag."\\ \hline
General & "I wonder why he beat her ass when she was tryna leave like she wasn't gone be running back when she found out she had HIV \& nobody want her"\\ \hline
General & "Aaannd. Hillary Clinton breathes a sigh of relief that Twitter has left its outrage of her AIDS comments behind to tend to Drumpf debacle."\\ \hline

@@

Another negative tweet questions whether Truvada is available as a generic drug. PrEP is currently only available through Gilead's Truvada, though a generic may be available as soon as 2017~\cite{truvadagenericblog}. Current patients must deal with uncertain drug prices, which vary from \$14,000 to \$70~\cite{truvadagenericblog}, and have in some cases, uncertain medical insurance reimbursement status. This tweet highlights both the difficulty of aquiring affordable PrEP medication, and also a level of conflicting information circulating on social media.

@@

TODO: investigate the reference to  \#truvadawhore which came up in the sentiment tweets above.

@@

- Moving towards HIV outbreak prediction.

- Automatically identifying PrEP adoption/adherence sentiment in US population.

- A lot of information on twitter is from bots, they only cost \$ per bot. This along with the character limit and slang makes twitter hard to use, but we've shown that it can be used. In fact twitter bots are slowly making twitter unusable because free advertising and marketing. Maybe in a future project we make twitter bots to spread information about PrEP.

>>>>

Pre-Exposure Prophylaxis (PrEP) is a recently developed therapy for the prevention of Human Immunodeficiency Virus (HIV) transmission. The treatment is available as a pill named Truvada and is taken once a day by HIV negative individuals for as much as 99\% protection when exposed to HIV. While medically effective, PrEP suffers from high drug costs, HIV-related social stigma, and under-informed health providers and patients. Data mining of social media has proven effective in the monitoring of HIV in the United States, but no study has investigated PrEP using social media. This paper describes a data mining and machine learning strategy using natural language processing (NLP), that monitors Twitter social media data to identify PrEP discussion trends. Our results show that we can identify PrEP and HIV discussion dynamics over time, and PrEP-related tweets with positive and negative sentiment. These results can be used by public health professionals to monitor PrEP discussion trends, and identify strategies to improve HIV prevention.

@@

Tweets from Twitter consist of short 140 character text messages that may also contain hashtag annotations. Twitter has been used as a source of data for large scale opinion mining in public health monitoring contexts to predict the spread of influenza~\cite{aramaki2011twitter}, predict postpartum depression~\cite{de2013predicting}, and examine tobacco use~\cite{myslin2013using}. Recently it has also been used for the study of HIV~\cite{young2014methods,young2013online}. These studies of HIV have focused on county-level HIV prevalence prediction, and general HIV discussion monitoring, but they have not focused on PrEP related discussion. Furthermore, existing HIV social media analyses have not taken full advantage of natural language processing (NLP) techniques to discover semantic information in unstructured text~\cite{young2015big}.

@@

Latent Dirichlet Allocation~\cite{blei2003latent} (LDA) is used to identify a small set of latent topics present in an unstructured corpus. LDA is a graphical model that generates documents from a set of latent topics. A topic is a probability distribution over words that captures a set of related words. LDA models are often inferred in practice using Bayesian inference either via collapsed Gibbs sampling or Variational Bayesian methods. Inspection of the resulting topics allows one to identify relevant terms and the context in which they occur in the corpus.

Dynamic Topic Modelling~\cite{blei2006dynamic} (DTM) is an extension of LDA that produces a series of topic models over time. Briefly, the documents in the corpus are divided into several corpuses that are successive in time. LDA is performed on each corpus to extract a topic distribution. The posterior topic distribution from time $t_n$ is used as the prior for time $t_{n+1}$. This lets the DTM model determine a topic model for each time point that is dependent both on the set of tweets from that time point, and on the previous time point's topic distribution.

@@

% table 1 - w2v PrEP
\begin{table}
\centering
\caption{Cosine similarity to word-vector "PrEP"}
\begin{tabular}{|l|c|} \hline
Related word & cosine similarity to PrEP\\ \hline
truvada & 0.796666\\ \hline
DoingIt & 0.738141\\ \hline
WorldAIDSDay & 0.720910\\ \hline
NancyReagan & 0.717667\\ \hline
NBHAAD & 0.705061\\ \hline
ART & 0.704300\\ \hline
HLM2016AIDS & 0.698698\\ \hline
ICASA2015 & 0.693117\\ \hline
HIV & 0.692860\\ \hline
OneConversation & 0.688427\\ \hline
\hline\end{tabular}
\end{table}

@@

We investigated the blog article linked to tweet 708519265540907010\cite{prephurtingcause}. The article, authored by Steven Banning, an HIV positive individual, expresses concerns about how new HIV treatments, like PrEP are being adopted by the HIV community. Steven notes that treatments for HIV positive individuals, and treatments to prevent HIV such as PrEP have made HIV a much more treatable disease than it was in the 1980's. These medical advances may have led in part to some unintended medical and social consequences. Steven notes that the rise of drug-resistant HIV strains has increased, in part due to HIV patients not adhering to to consistent treatment plans. He also points out the recent social trend of "bug chasers" individuals who are actively seeking to acquire and spread HIV strains. Perhaps the widespread availability of effective HIV medications have desensitized people to a healthy fear of the disease.

User 711275699529764864 appears to be a Twitter spam bot with no obvious connection to PrEP. Unfortunately, bots are commonly used on Twitter for advertising and marketing purposes, and sometimes hinder pertinent information retrieval. Thus while we observe some noise, our results demonstrate a way to quickly identify and monitor the most viral tweets related to PrEP. As the discussion changes, public health professionals can use this information to quickly identify the most relevant viral sentiment in the online PrEP conversation.

% table 2 - d2v PrEP
\begin{table}
\centering
\caption{Cosine similarity to document-vector "\#PrEP"}
\begin{tabular}{|l|c|} \hline
Related hashtag/tweet & cosine similarity to \#PrEP\\ \hline
\#lgbtmedia16 & 0.739128\\ \hline
\#hiv & 	0.727602 \\ \hline
\#whereisprep & 0.707165 \\ \hline
\#truvada & 0.696113 \\ \hline
\#hivprevention & 0.636068 \\ \hline
tweet-702179860983189504 & 0.630055\\ \hline
user-711275699529764864 & 0.629254\\ \hline
tweet-708519265540907010 & 0.628778 \\ \hline
tweet-712032637024653313 & 0.628646 \\ \hline
\#harrogatehour & 0.628547 \\ \hline
\hline\end{tabular}
\end{table}

@@

% do a top positive tweets table (also with tweet ID)
% table 3
\begin{table*}
\centering
\caption{Positive sentiment tweets.}
\begin{tabular}{|p{2.5cm}|p{12cm}|} \hline
Category & Text\\ \hline
General & "RT TOPublicHealth The Works provides testing for HIV anonymous \& rapid test available . Call 416-392-0520 for more info"\\ \hline
General & "RT FCAA ejaforg announced 5.4 million in grants to support orgs addressing \#HIV in new \& innovative ways!"\\ \hline
General & "RT HillaryClinton A note on the fight against HIV and AIDS and the people who really started the conversation."\\ \hline

@@

% do a top negative tweets table (also with tweet ID)
% table 4
\begin{table*}
\centering
\caption{Negative sentiment tweets.}
\begin{tabular}{|p{2.5cm}|p{12cm}|} \hline
Category & Text\\ \hline
General & "Also, how f***ing vile of Hillary to say. Reagan did f***ing NOTHING during the AIDS epidemic until it was too late. What a stupid old hag."\\ \hline
General & "I wonder why he beat her a** when she was tryna leave like she wasn't gone be running back when she found out she had HIV \& nobody want her"\\ \hline
General & "Aaannd. Hillary Clinton breathes a sigh of relief that Twitter has left its outrage of her AIDS comments behind to tend to Drumpf debacle."\\ \hline

@@

Another negative tweet questions whether Truvada is available as a generic drug. PrEP is currently only available through Gilead's Truvada, though a generic may be available as soon as 2017~\cite{truvadagenericblog}. Current patients must deal with uncertain drug prices, which vary from \$14,000 to \$70~\cite{truvadagenericblog}, and have in some cases, uncertain medical insurance reimbursement status. This tweet highlights both the difficulty of acquiring affordable PrEP medication, and also a level of conflicting information circulating on social media.

@@

The last tweet in the positive tweets table (Table 3), referencing the hashtag "\#truvadawhore", links to a blog article written by Evan Peterson titled "The Case for PrEP, or How I Learned to Stop Worrying and Love HIV-Positive Guys"~\cite{caseforprep}. Evan is an HIV-negative individual who adheres to a daily PrEP regimen to protect himself from HIV. Even finds that for him the side effects and inconveniences associated with taking daily Truvada are worth the protection, noting that he also practices safe sex, and gets tested regularly. Evan addresses the hashtag \#truvadawhore, which originated with a Huffington Post article titled "Truvada Whores?"~\cite{truvadawhore} which expressed concern toward the increased sexual risk taking of some Truvada patients. Even notes that for HIV-negative Truvada patients like him, who are careful about the risks they are taking, taking Truvada may actually provide more than just a medical protection. Truvada users may also be more informed than at risk people who do not use Truvada, because they are more educated and responsible with respect to their sexual health.

@@

In this article we describe a variety of NLP techniques that were used to analyse a tweet corpus to monitor the social media discussion for HIV prevention. We identified several notable trends. Firstly, we found that PrEP discussion is increasing on social media, Secondly we found that people who mention PrEP also talk about general health, STDs, stigma, and politics.

Our sentiment analysis identified both positive and negative PrEP and HIV related tweets. We identified two important arguments for an against PrEP. One as demonstrated by blog author Steven Banning hypothesises that PrEP overall could lead to unintended, and unhealthy risk seeking recklessness. The contrary hypothesis, described by blog author Even Peterson, suggests that PrEP adoption is a product of healthy, responsible at-risk individuals who are gaining HIV protection in an effective form. These opposing arguments were identified automatically using large scale social media data. Further investigation and data analysis, of social media data, or other health data may be able to quantify PrEP and Truvada's effectiveness in the ongoing effort to prevent HIV infection.

Future analyses may also take advantage of the rich Twitter API. An additional refinement of the analyses presented here may take a more active approach. Rather than simple passive data collection, we may build a twitter bot to interact with HIV and PrEP-tweeters, or conduct a distributed health survey. Finally, future HIV and PrEP research may benefit from the incorporation of multiple data sources, either social media sources, health information sources, or both.
